---
title: "Track Assignment"
subtitle: "Hallie Swan and Reid Shaw"
output:
  html_document: default
  pdf_document: default
date: "4/2/2019"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
```

100 students rank 16 tracks from 1-16, where 1 is the most desired track. Each student is assigned to one track. Each track can hold 8 students. Some tracks are more popular than others. How do we optimize assignment of students to tracks, such that we minimize the total rank score (i.e. more students have lower rank scores and are in more highly desired tracks)?  

Currently, Rochester iteratively selects a random student and assigns them to their highest rank track that is available (Or that is my best guess as to what they do). This can result in some students being assigned to poorly ranked tracks. Here, we explore whether solving the assignment problem using the Hungarian method results in overall minimized total rank score.

## Creating a Dataset of Randomized Rankings

We can generate a dataset containing randomly assigned ranks. We will weight the first four tracks more heavily to simulate the popularity of some tracks. The remaining tracks will be weighted evenly.

```{r create_randomized_ranks}
#' Create Randomized Track Ranks
#' 
#' This function creates a data.frame of students with randomized ranks 
#' of various tracks, where a smaller integer indicates a stronger preference for 
#' the given track. The first four tracks will be weighted more heavily (i.e. 
#' are more popular choices). The remaining tracks will be weighted evenly.
#'
#' @param seed an integer
#' @param n_students an integer: the number of students
#' @param n_tracks an integer: the number of tracks
#'
#' @return data.frame with columns track, student, rank
#' @export
#'
#' @examples
#' create_random_ranks(125)
create_random_ranks <- function(seed, n_students = 100, n_tracks = 16) {
    set.seed(seed)
    ranks <- unlist(lapply(1:n_students, function(x) {
        sample.int(n_tracks, n_tracks, replace = FALSE, 
                   prob = c(0.25, 0.25, 0.1, 0.2, 
                            rep(0.2/(n_tracks-4), times = (n_tracks-4))))
    }))
    tracks <- rep(sprintf("track%02d", 1:n_tracks), times = n_students)
    students <- rep(sprintf("student%03d", 1:n_students), each = n_tracks)
    data.frame(
        track = tracks,
        student = students,
        rank = ranks,
        stringsAsFactors = FALSE
    )
}
```

We can plot an example dataset to ensure that there is a good distribution of track ranking. A few rows of the dataset are shown below:

```{r example_dataset}
df <- create_random_ranks(25)
head(df)
```

We plot the tracks by the number of students giving a certain rank. We see that the first four tracks have the highest ranks, but the remaining tracks have a good distribution of rankings.

```{r plot_example_dataset, fig.height=12, fig.width=11}
df %>%
    group_by(rank, track) %>%
    count() %>%
    ggplot(aes(x = rank, y = n)) +
    facet_wrap(~track) +
    geom_bar(stat = "identity") +
    labs(title = "Number of Students Giving a Rank Per Track",
                  y = "Number of Students",
                  x = "Rank")
```

## Using the Rochester Method

We can then assign students to tracks using the Rochester method. 

```{r use_rochester_method}
#' Use Rochester Method to Assign Students to Tracks
#' 
#' This function will take a data.frame of student ranks per track and 
#' will assign students to tracks using the Rochester method. The Rochester 
#' method iteratively selects a random student and assigns them to their 
#' highest rank track that is available. The goal is to minimize each 
#' student's ranks, i.e. a rank of 1 should be the student's 
#' top choice track. 
#'
#' @param df data.frame with three columns: track, student, rank
#' @param starting_seed integer
#' @importFrom dplyr group_by ungroup filter bind_rows count pull
#' @importFrom tidyr unite
#'
#' @return
#' @export
#'
#' @examples
#' df <- create_random_ranks(25)
#' use_rochester_method(df)
use_rochester_method <- function(df, n_seats = 8, starting_seed = 1234) {
    
    # generate random student order
    set.seed(starting_seed)
    student_order <- sample(unique(df$student))
    
    # assign first student
    assignments <- df %>% 
        dplyr::filter(student == student_order[1] & rank == 1)
    
    # for remaining students
    # remove full tracks
    # assign student to highest rank available track
    for(i in student_order[-1]) {
        full_tracks <- assignments %>%
            dplyr::group_by(track) %>%
            dplyr::count() %>%
            dplyr::filter(n == n_seats) %>%
            dplyr::pull(track)
        assignments <- df %>%
            dplyr::filter(!(track %in% full_tracks)) %>%
            dplyr::filter(student == i) %>%
            dplyr::filter(rank == min(rank)) %>%
            dplyr::bind_rows(assignments)
    }
    # return assignments
    # add seat number
    assignments %>%
        dplyr::group_by(track) %>%
        dplyr::mutate(seat = sprintf("seat%02d", row_number())) %>%
        tidyr::unite(track_seat, track, seat, sep = "_") %>%
        dplyr::ungroup() %>%
        as.data.frame()
}
```

Using our example dataset, we look at the rankings of the tracks to which students were assigned using the Rochester method.

```{r rochester_plot}
rochester <- use_rochester_method(df)
rochester %>%
    group_by(rank) %>%
    count(name = "n_students") %>%
    ungroup() %>%
    mutate(rank = factor(rank, levels = 1:16)) %>%
    ggplot(aes(x = rank, y = n)) +
    geom_bar(stat = "identity") +
    scale_x_discrete(drop=FALSE) +
    labs(title = "Rochester Method Assignment Rankings")
```

We see that most students were assigned to a track that they had ranked in the top half of tracks. We can think of the total "cost" of the assignment to be the sum of the rankings for each student. The total cost for this Rochester assignment is `r sum(rochester$rank)`.

## Maximizing First Choices Method

This method is similar to Rochester method, but it always has the highest number of first choices

```{r use_max_first}
use_max_first <- function(df, n_seats = 8, starting_seed = 1234) {
     
     # generate random student order
     set.seed(starting_seed)
     student_order <- sample(unique(df$student))
     
     # assign first student
     assignments <- df %>% 
          dplyr::filter(student == student_order[1] & rank == 1)
     
     # for remaining students
     # remove full tracks
     for(k in sort(unique(df$rank))){
               for (j in student_order[-1]) {
                    full_tracks <- assignments %>%
                         dplyr::group_by(track) %>%
                         dplyr::count() %>%
                         dplyr::filter(n == n_seats) %>%
                         dplyr::pull(track)
                    used_students <- assignments %>%
                         dplyr::group_by(student) %>%
                         dplyr::pull(student)
                    assignments <- df %>%
                         dplyr::filter(!(track %in% full_tracks)) %>%
                         dplyr::filter(!(student %in% used_students)) %>%
                         dplyr::filter(student == j) %>%
                         dplyr::filter(rank == k) %>%
                         dplyr::bind_rows(assignments)
          }
     }
     # return assignments
     # add seat number
     assignments %>%
          dplyr::group_by(track) %>%
          dplyr::mutate(seat = sprintf("seat%02d", row_number())) %>%
          tidyr::unite(track_seat, track, seat, sep = "_") %>%
          dplyr::ungroup() %>%
          as.data.frame()
}
```

Here, we look at the rankings of the tracks to which students were assigned using the maximizing 1st choices method.

```{r}
first <- use_max_first(df)
first %>%
     group_by(rank) %>%
     count(name = "n_students") %>%
     ungroup() %>%
     mutate(rank = factor(rank, levels = 1:16)) %>%
     ggplot(aes(x = rank, y = n)) +
     geom_bar(stat = "identity") +
     scale_x_discrete(drop=FALSE) +
     labs(title = "Maximizing First Preferences Method Assignment Rankings")
```

## Using the Hungarian Method

We can then assign students using the [Hungarian Method](http://www.hungarianalgorithm.com). We utilize the `solve_LSAP` function from the **clue** package to perform the algorithm. The Hungarian method also requires a square matrix of values, so we will create "dummy students" to fill out all available seats in each track. We will give each track the worst ranking so that these dummy students don't affect the assignment of our real students.

```{r use_hungarian_method}
#' Use Hungarian Method to Assign Students to Tracks
#' 
#' This function will take a data.frame of student ranks per track and 
#' will assign students to tracks using the Hungarian method. Per documentation 
#' for clue::solve_LSAP, the Hungarian method finds an optimal assignment 
#' of rows to columns using primal-dual methods. The goal is to minimize each 
#' student's ranks, i.e. a rank of 1 should be the student's 
#' top choice track. 
#'
#' @param df data.frame with three columns: track, student, rank
#' @param n_seats integer: the number of seats per track
#' @importFrom dplyr n_distinct bind_rows filter
#' @importFrom tidyr spread
#' @importFrom clue solve_LSAP
#'
#' @return data.frame with columns: track_seat, student, rank
#' @export
#'
#' @examples
#' df <- create_random_ranks(25)
#' use_hungarian_method(df)
use_hungarian_method <- function(df, n_seats = 8) {
    n_tracks <- dplyr::n_distinct(df$track)
    n_students <- dplyr::n_distinct(df$student)
    # create dummy data - assign highest rank
    # total number of students for square matrix - number of students present
    total_students <- n_tracks*n_seats
    # calculate number of dummy students
    n_dummy <- total_students - n_students
    dummy_ranks <- rep(max(df$rank), times = n_dummy*n_tracks)
    dummy_tracks <- rep(sprintf("track%02d", 1:n_tracks), times = n_dummy)
    dummy_students <- rep(sprintf("student%03d", (n_students+1):total_students), 
                          each = n_tracks)
    dummy <- data.frame(
        track = dummy_tracks,
        student = dummy_students,
        rank = dummy_ranks,
        stringsAsFactors = FALSE
    )
    df <- dplyr::bind_rows(df, dummy)
    # replicate data per seat
    df <- lapply(1:n_seats, function(x) {
        df$track_seat <- paste(df$track, sprintf("%02d", x), sep = "_seat")
        df
    }) %>%
        dplyr::bind_rows() %>%
        dplyr::select(-track) %>%
        tidyr::spread(student, rank)
    # make into matrix
    mat <- as.matrix(df[, -1])
    # check that square
    if(nrow(mat) != ncol(mat)) {
        stop("Matrix needs to be square")
    }
    # use hungarian method
    solved <- clue::solve_LSAP(mat, maximum = FALSE)
    # generate assignments
    assignments <- data.frame(
        track_seat = df$track_seat,
        student = unlist(lapply(seq_along(solved), function(x) colnames(mat)[solved[x]])), 
        rank = mat[cbind(seq_along(solved), solved)],
        stringsAsFactors = FALSE
    )
    # remove dummy students and return assignments
    assignments %>%
        dplyr::filter(!(student %in% dummy_students))
}
```

Using the same example dataset, we look at the rankings of the tracks to which students were assigned using the Hungarian method.

```{r hungarian_plot}
hungarian <- use_hungarian_method(df)
hungarian %>%
    group_by(rank) %>%
    count(name = "n_students") %>%
    ungroup() %>%
    mutate(rank = factor(rank, levels = 1:16)) %>%
    ggplot(aes(x = rank, y = n)) +
    geom_bar(stat = "identity") +
    scale_x_discrete(drop=FALSE) +
    labs(title = "Hungarian Method Assignment Rankings")
```

We see that more students were assigned to a higher priority track. The total cost for this Hungarian assignment is `r sum(hungarian$rank)` (if everyone got their first choice the cost would be 100). This is lower than the Rochester method assignment cost. 


## Iterate Methods

We can now compare the methods across many example simulated datasets to see if either method consistently returns a lower total assignment cost.  

```{r iterate_methods}
#' Iterate Rochester and Hungarian Method To Compare Cost
#' 
#' This function will iteratively perform student track assignment from datasets 
#' with random ranking of tracks using the Rochester and Hungarian methods. The 
#' total ranking "cost" will be returned for each method. A lower cost indicates 
#' that more students received more desirable tracks.
#'
#' @param n_iterations integer, number of random ranking data.frames to compare
#' @inheritParams use_rochester_method
#' @inheritParams use_hungarian_method
#' @inheritParams create_random_ranks
#'
#' @return
#' @export
#'
#' @examples
#' iterate_methods(10)
iterate_methods <- function(n_iterations, 
                            n_students = 100, n_tracks = 16, n_seats = 8, 
                            starting_seed = 1234) {
    lapply(1:n_iterations, function(i) {
        df <- create_random_ranks(i, n_students, n_tracks)
        hungarian <- use_hungarian_method(df, n_seats)$rank
        hungarian <- data.frame(
            cost = sum(hungarian),
            max_rank = max(hungarian),
            min_rank = min(hungarian),
            first_preference = sum(hungarian == 1)
        )
        rochester <- use_rochester_method(df, n_seats, starting_seed)$rank
        rochester <- data.frame(
            cost = sum(rochester),
            max_rank = max(rochester),
            min_rank = min(rochester), 
            first_preference = sum(rochester == 1)
        )
        first <- use_max_first(df, n_seats, starting_seed)$rank
        first <- data.frame(
            cost = sum(first),
            max_rank = max(first),
            min_rank = min(first), 
            first_preference = sum(first == 1)
        )
        data.frame(
            iteration = i,
            method = c("hungarian", "rochester", "first"),
            cost = c(hungarian$cost, rochester$cost, first$cost),
            max_rank = c(hungarian$max_rank, rochester$max_rank, first$max_rank),
            min_rank = c(hungarian$min_rank, rochester$min_rank, first$min_rank),
            first_preference = c(hungarian$first_preference, rochester$first_preference, first$first_preference),
            stringsAsFactors = FALSE
        )
    }) %>%
        bind_rows()
}
```

Here, we compare 1000 iterations of the example dataset. We first plot the total cost of the methods across each iteration. 

```{r plot_cost_comparison}
compare_methods <- iterate_methods(1000)
compare_methods %>%
    ggplot(aes(x = iteration, y = cost, group = method, color = method)) +
    geom_line() +
    labs(title = "Cost Per Iteration") +
    coord_cartesian(ylim = c(100, 400))
```

We can then plot the worst rank in each iteration.

```{r}
compare_methods %>%
    ggplot(aes(x = iteration, y = max_rank, group = method, color = method)) +
    geom_line() +
    labs(title = "Worst Track Rank Per Iteration") +
    coord_cartesian(ylim = c(0, 15))
```

We can also plot the number of times a student got their first choice in each iteration.

```{r}
compare_methods %>%
    ggplot(aes(x = iteration, y = first_preference, group = method, color = method)) +
    geom_line() +
    labs(title = "Number of First Preferences Achieved Per Iteration") +
    coord_cartesian(ylim = c(0, 100))
```

We see that in nearly all iterations, the Hungarian method has a lower cost, and lower worst rank. Unsurprisingly, iteratively assigning students to their top choice consistently has the highest number of first preferences achieved. The Rochester method is not the best in any of these categories. This means that the Hungarian method completes the assignment problem more effectively than the other two methods, especially the Rochester method. 

## Summary

Based on this study, it appears that the Hungarian method is more effective at minimizing cost than the Rochester method. Therefore, the Hungarian method would better assign students to tracks than the Rochester method. 
